---
title: "Garmin Demo"
output: 
  flexdashboard::flex_dashboard:
    orientation: rows
    #social: menu
    #source_code: embed
    vertical_layout: fill
---

```{r setup}
library(flexdashboard)
library(htmlwidgets)   ## data viz
library(plotly)        ## data viz
library(dygraphs)      ## data viz
library(leaflet)       ## data viz
library(ggplot2)       ## data viz
library(lubridate)     ## manipulating dates data
library(changepoint)   ## changepoint
library(Benchmarking)  ## efficient frontier analysis
```

Introduction
=======================================================================

I'm a runner and a data junkie, therefore I love my [Garmin Forerunner 620](https://buy.garmin.com/en-US/US/wearabletech/wearables/previous_models_wearables/forerunner-620/prod122785.html). However, when I can't find the insights I want in [Garmin Connect](http://connect.garmin.com), I do my own statistical modeling. This portfolio demonstrates some of the advanced analytics I've developed using my personal running and cycling data. Each tab to the right includes self-contained analyses:

* Efficient Frontier - compares activities from recent history and evaluates how efficient each run is to the aggregate
* Changepoint Analysis - evaluates changes in performance during recorded activities
* More - trend analysis reviews historical activity metrics (pace, heart rate, VO2 max, etc) and suggests how they may evolve moving forward

All mathematical modeling was performed using the R software package. Source code can be found via the following Github repository: [https://github.com/johnsug/garmin](https://github.com/johnsug/garmin)

Changepoint Analysis
=======================================================================

Row {data-height=50}
-----------------------------------------------------------------------

Changepoint analysis is super cool, yo.

Row
-----------------------------------------------------------------------

### Event 1

```{r}

```

Row
-----------------------------------------------------------------------

### Event 2

```{r}

```


Efficient Frontier
=======================================================================

Row {data-height=50}
-----------------------------------------------------------------------

Efficient frontiers are cool, too.


Row
-----------------------------------------------------------------------

### Efficient Frontier

```{r}

```


Other Ideas
=======================================================================

Row
-----------------------------------------------------------------------

### Chart A

```{r}

```

Row
-----------------------------------------------------------------------

### Chart B

```{r}

```
