---
title: "Garmin Demo"
output: 
  flexdashboard::flex_dashboard:
    orientation: rows
    vertical_layout: scroll
    source_code: embed
runtime: shiny
---

```{r fix list}
# optimize for mobile viewing ???
```

```{r setup}
library(flexdashboard)    ## dashboard framework
library(shinydashboard)   ## more dashboard framework
library(shiny)            ## dashboard widgets
library(htmlwidgets)      ## data viz
library(plotly)           ## data viz
library(ggplot2)          ## data viz
library(dygraphs)         ## data viz (plot used in changepoint study)
library(googleVis)        ## data viz (calendar heatmap)
library(changepoint)      ## changepoint analysis
library(data.table)       ## data munging
library(plyr)             ## data munging
library(formattable)      ## formatting
library(lubridate)        ## manipulating dates data
library(xts)              ## extended time series functionality

# shortcut wrappers
DF <- function(...) { data.frame(...) }
DT <- function(...) { data.table(...) }

# miles per hour (MPH) to pace custom function
mph_to_pace <- function(mph){
  pace <- 60/mph
  pace_mins <- floor(pace)
  pace_secs <- round((pace - pace_mins)*60)
  pace_spacer <- ":"
  if(nchar(pace_secs)==1) {pace_spacer <- ":0"}
  pace <- paste0(pace_mins, pace_spacer, pace_secs)
  return(pace)
}

# miles per hour (MPH) to pace custom function
pace_to_mph <- function(pace){
  mph <- as.numeric(gsub(":", ".", pace))
  mph_full <- floor(mph)
  mph_part <- (mph - mph_full)/.60
  pace <- 60 / (mph_full + mph_part)
  return(round(pace,2))
}

# pace to raw pace
raw_pace <- function(pace){
  raw <- DT(t(DF(strsplit(pace, ":"))))
  raw <- as.numeric(raw$V1) + as.numeric(raw$V2)/60
  return(raw)
}

## load datasets
# daily log (agregated)
d <- fread("https://raw.githubusercontent.com/johnsug/garmin/master/forerunner_log.csv")
d <- d[!is.na(Distance)]
names(d) <- gsub(" ", "_", names(d))
names(d)[names(d)=="Avg_Speed(Avg_Pace)"] <- "Pace"
d$MPH <- pace_to_mph(d$Pace)
d$Date <- as.Date(substr(d$Start,6,100), format="%b %d, %Y")
d$Mins <- 1/d$MPH * d$Distance * 60
d$Raw_Pace <- raw_pace(d$Pace)
dat <- DT(d)[Activity_Type=="Running"][Max_HR!="--"]
dat$Full_Activity_Name <- paste(dat$Activity_Name, "-", dat$Start)

# detailed logs
f <- fread("https://raw.githubusercontent.com/johnsug/garmin/master/detailed_forerunner_log.csv")
f$Time_Stamp <- paste(f$Date, f$Time)
f$Time_Stamp <- as.POSIXct(f$Time_Stamp)
tz(f$Time_Stamp) <- "America/Chicago"
f$DistanceMeters <- as.numeric(f$DistanceMeters)
f$HeartRateBpm <- as.numeric(f$HeartRateBpm)

# personal records
pr <- DF(fread("https://raw.githubusercontent.com/johnsug/garmin/master/personal_records.csv"))
is_hours <- substr(pr$Time,2,2)==":"
pr$Time <- strsplit(pr$Time, ":")
for(i in 1:nrow(pr)){
  if(is_hours[i]==F) {pr$Time[i] <- as.numeric(pr$Time[[i]][1]) + 
                                 as.numeric(pr$Time[[i]][2])/60}
  
  if(is_hours[i]==T) {pr$Time[i] <- as.numeric(pr$Time[[i]][1])*60 + 
                                    as.numeric(pr$Time[[i]][2]) + 
                                    as.numeric(pr$Time[[i]][3])/60
  }
}; rm(i)
pr$Time <- as.numeric(pr$Time)

a <- d[Activity_Type=="Running"][, .(Date=as.Date(Date, format="%m/%d/%Y"), 
                                     Distance=Distance, 
                                     Time=Time, 
                                     Average_Heart_Rate=as.numeric(Max_HR), 
                                     Average_Pace=raw_pace(Pace), 
                                     Miles_Per_Hour=MPH, 
                                     Training_Effect=as.numeric(Training_Effect), 
                                     V02_Max=V02_Max)]
names(a) <- gsub("_", " ", names(a))

# date vectors
today <- Sys.Date()
dates_2_months <- today-60
dates_6_months <- today-180
dates_12_months <- today-365
```



Introduction
=======================================================================

<h1><i class="fa fa-bar-chart"></i> <b>Digital Portfolio</b></h1>

<h4>I am a runner and a data junkie, which means I love reviewing my personal fitness activities recorded by my [Garmin Forerunner 620](https://buy.garmin.com/en-US/US/wearabletech/wearables/previous_models_wearables/forerunner-620/prod122785.html) GPS watch and heart rate monitor. The [Garmin Connect](http://connect.garmin.com) dashboards that sync with my watch are great; however when I can't find the insights I want, I perform my own statistical modeling using my personal running data. This dashboard showcases some of the analytical views I've developed, to better track my own running activity:
  
* <a href=#trend-analysis>Trend Analysis</a> - reviews historical activity metrics (heart rate, pace, V0<sub>2</sub> max, etc)
* <a href=#pace-benchmarking>Pace Benchmarking</a> - compares recent runs to historical averages and records
* <a href=#calendar-heatmap>Calendar Heatmap</a> - quickly illustrates annual activity level
* <a href=#changepoint-analysis>Changepoint Analysis</a> - drills down on changes in intensity during individual activities

<br>
This digital portfolio was created by John Sugden using the 'flexdashboard' package in R. For more about John, please visit my <a href=http://johnsug.github.io><i class="fa fa-rss-square"></i> blog</a> or my <a href=https://www.linkedin.com/in/johnsugden><i class="fa fa-linkedin-square"></i> LinkedIn</a> profile.</h4>




Trend Analysis
=======================================================================

```{r trend}
activity_types <- sort(names(a)[-(1:3)])

shinyApp(
  ui = fillPage(
    fluidRow(
      box(uiOutput("trend_horizon"), width=4), 
      box(selectInput("trend_box_1", "Plot 1 metric:", choices=activity_types, selected=activity_types[1]), width=3), 
      box(selectInput("trend_box_2", "Plot 2 metric:", choices=activity_types, selected=activity_types[2]), width=3)), 
    
    fluidRow(
      box(
        plotlyOutput("trend_plot_1", height="275px"), 
        HTML("<center><h4><font color=0000FF>"), textOutput("trend_text_1"), HTML("</center></h4></font>"), 
        width=6), 
    
      box(
        plotlyOutput("trend_plot_2", height="275px"), 
        HTML("<center><h4><font color=0000FF>"), textOutput("trend_text_2"), HTML("</center></h4></font>"), 
        width=6)
    )),
  
  server = function(input, output) {
    
    ## trend horizon adaptive radio buttons
    output$trend_horizon <- renderUI({
      pace_check <- sum(!is.na(a$`Miles Per Hour`[a$Date > dates_2_months]))
      hr_check <- sum(!is.na(a$`Average Heart Rate`[a$Date > dates_2_months]))
      vo2_check <- sum(!is.na(a$V02[a$Date > dates_2_months])) 
      
      ## hide if less than 4 data points
      if(pace_check < 4 || hr_check < 4 || hr_check < 4) { 
        flex_list <- list("6 months"=6, "12 months"=12) 
        }
      else { 
        flex_list <- list("2 months"=2, "6 months"=6, "12 months"=12) 
        }
      
      ## return UI widget
      radioButtons("trend_horizon", label="Time horizon for plots:", choices=flex_list, inline=T, selected=12)
    })
    
    ## filter data based on radio buttons
    plot_dat <- reactive({
      pd <- a
      if(input$trend_horizon==2) {pd <- pd[Date > dates_2_months]}
      if(input$trend_horizon==6) {pd <- pd[Date > dates_6_months]}
      if(input$trend_horizon==12) {pd <- pd[Date > dates_12_months]}
      pd$Pace <- NULL
      pd <- melt(pd, id.vars=c("Date", "Distance", "Time"), variable.name="Plot", value.name="Value")
      pd
    })
    
    ## period trend change table
    change_table <- reactive({
      pd <- plot_dat()
      metrics <- DT(pd)[!is.na(Value), .(Start=head(Value,1), End=tail(Value,1)), by=Plot]
      metrics$Change <- metrics$Start - metrics$End
      metrics$Percent_Change <- round(metrics$Start / metrics$End - 1,3)
      metrics$Periodic_Change <- scales::percent(metrics$Percent_Change)
      metrics[, .(Plot=Plot, Periodic_Change=Periodic_Change)]
    })
    
    ## plot #1: build ggplot then wrap with ggplotly
    gg_01 <- reactive({
      pd <- subset(plot_dat(), Plot==input$trend_box_1)
      g1 <- ggplot(pd, aes(x=Date, y=Value)) + 
              geom_line(col="gray") + 
              geom_point(size=2) + 
              ## lm, glm, gam, rlm, loess, 
              stat_smooth(method="lm", col="blue", fill="royalblue", alpha=.3, formula=y ~ x + I(x^2)) + 
              scale_x_date(date_breaks="2 months", date_labels="%b %d")
      
      ggplotly(g1)
    })
    
    ## plot #2: build ggplot then wrap with ggplotly
    gg_02 <- reactive({
      pd <- subset(plot_dat(), Plot==input$trend_box_2)
      g2 <- ggplot(pd, aes(x=Date, y=Value)) + 
              geom_line(col="gray") + 
              geom_point(size=2) + 
              ## lm, glm, gam, rlm, loess, 
              stat_smooth(method="lm", col="blue", fill="royalblue", alpha=.3, formula=y ~ x + I(x^2)) + 
              scale_x_date(date_breaks="2 months", date_labels="%b %d")
      
      ggplotly(g2)
    })
  
    ## render text
    output$trend_text_1 <- renderText({
      paste("Periodic change =", DT(change_table())[Plot==input$trend_box_1, Periodic_Change])
    })
    
    output$trend_text_2 <- renderText({
      paste("Periodic change =", DT(change_table())[Plot==input$trend_box_2, Periodic_Change])
    })
    
    ## render plots
    output$trend_plot_1 <- renderPlotly(
      gg_01() %>% config(displayModeBar=F)
    )
    
    output$trend_plot_2 <- renderPlotly(
      gg_02() %>% config(displayModeBar=F)
    )
  }
)
```

<h4><a href=#changepoint-analysis><i class="fa fa-arrow-left"></i></a> <a href=#pace-benchmarking><i class="fa fa-arrow-right"></i></a><br>

This widget shows how one's fitness metrics are changing over a selected time horizon by way of trend analysis. The user has the ability to select from five metrics: average heart rate, average pace, average speed (MPH), training effect, and V0<sub>2</sub> max. After selecting a metric from the drop-down menus, the user's historical metric activity is plotted (black dots) and connected with a simple gray line. 

Linear regression methods are used to fit a trend line (dark blue line) to the data points. The light blue shading indicates the 95% confidence interval around the fitted trend line. The trend line suggests which direction the user's metrics are trending in the near future.

Lastly, the percent change between the first and last activities in the selected time horizon is provided as the periodic change. </h4>




Pace Benchmarking
=======================================================================

```{r benchmarking}

shinyApp(
  ui = 
    fillPage(
      
      ## control widgets
      fluidRow(box(radioButtons("points_horizon", 
                            label="Time horizon for plot:", 
                            choices=list("2 months"=2, "6 months"=6, "12 months"=12, "All Time"=999), 
                            inline=T, 
                            selected=12), 
                   width=6)),
      
      fluidRow(box(plotlyOutput("mc_plot", height="300px"), width=12))
  ),
  
  server = function(input, output) {
    
    ## query data
    eff_dat <- reactive({
      
      ## call data
      eff_dat2 <- dat
      
      ## filter based on desired experience period
      if(input$points_horizon==2)  { eff_dat2 <- subset(eff_dat2, Date >= dates_2_months)  }
      if(input$points_horizon==6)  { eff_dat2 <- subset(eff_dat2, Date >= dates_6_months)  }
      if(input$points_horizon==12) { eff_dat2 <- subset(eff_dat2, Date >= dates_12_months) }
      
      ## return data frame
      eff_dat2
    })
    
    ## most recent activity
    recent <- head(dat, 1)
    
    ## build "best pace" frontier using personal records file
    best_pace <- DT(pr)[, .(x=Distance, y=Time)]
    slope <- lm(tail(best_pace$y,2)~tail(best_pace$x,2))$coefficients[2]
    if(is.na(slope)) { slope <- best_pace$y/best_pace$x}
    inc <- (ceiling(max(best_pace$x)) - max(best_pace$x))*slope
    best_pace <- rbind(DF(x=0, y=0), best_pace, DF(x=ceiling(max(dat$Distance)), y=max(best_pace$y)+inc))
    
    ## fit average pace OLS regression
    avg_pace <- reactive({
      avg_dat <- eff_dat()
      avg <- lm(Mins~Distance, dat=avg_dat)
      upper_bound <- max(best_pace$x)
      avg_frontier <- rbind(DF(x=0, y=avg$coefficients[1]), 
                            DF(x=upper_bound, 
                               y=max(avg$coefficients[1] + upper_bound*avg$coefficients[2])))
      return(avg_frontier)
    })
    
    ## best time / average time plot
    output$mc_plot <- renderPlotly({
      
      ## plot points
      plot_ly(eff_dat(), x=Distance, y=Mins, name="Individual Runs", hoverinfo="text", 
              text=paste0(Start, 
                          "<br>Distance: ", Distance, 
                          "<br>Elapsed Time: ", Time, 
                          "<br>Average Pace: ", Pace), 
              mode="markers", marker=list(size=10), opacity=.75, showlegend=T) %>%
        config(displayModeBar=F) %>% 
        layout(xaxis=list(title="Distance (Miles)", fixedrange=TRUE), 
               yaxis=list(title="Run Time (Minutes)", fixedrange=TRUE)) %>%
      
        ## most recent activity
        add_trace(x=recent$Distance, y=recent$Mins, name="Most Recent", mode="markers", opacity=.5, 
                  marker=list(symbol="x", size=16, color="red"))  
            
        ## plot average-pace line
        add_trace(x=avg_pace()$x, y=avg_pace()$y, name="Average Pace", mode="lines", opacity=.3, 
                  line=list(width=5, color="cyan")) %>%
        
        ## plot best-pace line
        add_trace(x=best_pace$x, y=best_pace$y, name="Best Pace", mode="lines", opacity=.3, 
                  line=list(width=5, color="orange"))
          
      })
    }
)

```

<h4><a href=#trend-analysis><i class="fa fa-arrow-left"></i></a> <a href=#calendar-heatmap><i class="fa fa-arrow-right"></i></a><br>

In this plot, historical activities (blue dots) are compared to benchmarks of both historical best paces (gold line) and average paces (light blue line) from the selected time horizon.</h4>




Calendar Heatmap
=======================================================================

```{r heatmap}
shinyApp(
  ui = fillPage(htmlOutput("heatmap")),
 
  server = function(input, output) {
    output$heatmap <- renderGvis({
      gvisCalendar(data=subset(a, Date > "2015-12-31"), 
                   datevar="Date", 
                   numvar="Distance", 
                   chartid="Calendar", 
                   options=list(width=1000, height=200, title="Annual Activity (in Miles)", 
                                colorAxis="{minValue: 0, colors: ['yellow', 'green']}")
      )
    })
  }
)
```

<h4><a href=#pace-benchmarking><i class="fa fa-arrow-left"></i></a> <a href=#changepoint-analysis><i class="fa fa-arrow-right"></i></a><br>

A heatmap is a graphical representation of data values which illustrates high/low values, by way of color, without explicitly displaying the numeric values. 

This calendar heatmap displays both which days of the week/year I've run and how far I've run each day. (Yellow blocks indicate low mileage days, green blocks indicate average mileage days, and dark-green blocks indicate high mileage days). This heatmap is especially useful to hone in on day-of-the-week and week-to-week activity trends.</h4>




Changepoint Analysis
=======================================================================

```{r changepoint}
###http://rmarkdown.rstudio.com/flexdashboard/shiny.html
## I'm fighting limiting this to five segments, because I want to hold onto the colors.... perhaps I should just let it go, use PELT, and let the user determine the minimum segment length...

shinyApp(
  
  ## user interface #####
  ui = 
    fillPage(
      selectInput("changepoint_list", "Choose Activity", width=400, 
                  choices=dat[Tagging!="" & Tagging!="--" & !is.na(Tagging), Full_Activity_Name]), 
                    
      fluidRow(
        box(dygraphOutput("changepoint_plot", height="300px"), width=8),
        box(formattableOutput("changepoint_table"), width=4))
  ),
  
  ## server #####
  server = function(input, output) {
    
    ## subset changepoint data from full detailed log
      changepoint_data <- reactive({
      
      # filter data based on selection from drop-down menu
      cp_dat <- f[Activity==dat$Tagging[dat$Full_Activity_Name==input$changepoint_list]]
      
      # # set first point of time series to 0;00:01, to match up with race elapsed time
      # zero_out <- substr(cp_dat$Time_Stamp[1],12,19)
      # zero_out <- as.numeric(substr(zero_out,1,2))*3600 + as.numeric(substr(zero_out,4,5))*60 + 
      #   as.numeric(substr(zero_out,7,8)) - 1
      # cp_dat$Time_Stamp <- cp_dat$Time_Stamp - zero_out
      
      # return data frame
      cp_dat
    })
    
    ## segment time series (for plotting)
    time_series <- reactive({
      ts_dat <- changepoint_data()
      xts(x=ts_dat$MPH, order.by=ts_dat$Time_Stamp)
    })
    
    ## perform changepoint analysis
    cp <- reactive({
      
      ## find changepoints
      cp_dat <- changepoint_data()
      cp <- cpt.mean(cp_dat$MPH, method="BinSeg", Q=4, minseglen=10, penalty="AIC")
      cp <- c(1,cp@cpts)
      
      ## add redundant changepoints, if necessary
      if (length(cp) < 6) {cp <- c(cp, rep(max(cp), 6-length(cp)))}
      
      ## return changepoints
      cp
    })
    
    ## changepoint summary table
    cp_table_data <- reactive({
      cpts <- cp()
      cp_data <- changepoint_data()
      td <- DF(Zone=1:5, Minutes=0, Distance_Miles=0, MPH=0, Avg_Pace=0, Avg_HeartRate=0)
      for(i in 1:5){
        td$Minutes[i] <- round((cp_data$Elapsed_Seconds[cpts[i+1]] - 
                                  cp_data$Elapsed_Seconds[cpts[i]])/60,1)
        td$Distance_Miles[i] <- round((cp_data$DistanceMeters[cpts[i+1]] - 
                                         cp_data$DistanceMeters[cpts[i]]) * 0.000621371, 2)
        td$MPH[i] <- mean(cp_data$MPH[cpts[i]:(cpts[i+1]-1)])
        td$Avg_Pace[i] <- mph_to_pace(td$MPH[i])
        td$MPH[i] <- round(td$MPH[i],1)
        td$Avg_HeartRate[i] <- round(mean(cp_data$HeartRateBpm[cpts[i]:(cpts[i+1]-1)]))
      }
      names(td) <- c("Zone", "Time (Minutes)", "Distance (Miles)", "Average Speed (MPH)", "Average Pace", 
                     "Average Heart Rate")
      td[td$`Time (Minutes)`!=0,]
    })
    
    ## mean value time series
    mean_series <- reactive({
      ts_dat <- changepoint_data()
      cpts <- cp()
      cp_diff <- diff(cpts)
      m1 <- mean(ts_dat$MPH[cpts[1]:(cpts[2])])
      m2 <- mean(ts_dat$MPH[cpts[2]:(cpts[3]-1)])
      m3 <- mean(ts_dat$MPH[cpts[3]:(cpts[4]-1)])
      m4 <- mean(ts_dat$MPH[cpts[4]:(cpts[5]-1)])
      m5 <- mean(ts_dat$MPH[cpts[5]:(cpts[6]-1)])
      
      ## return mean values
      c(rep(m1, cpts[2]), 
        rep(m2, cpts[3]-cpts[2]), 
        rep(m3, cpts[4]-cpts[3]), 
        rep(m4, cpts[5]-cpts[4]), 
        rep(m5, cpts[6]-cpts[5]))
    })
    
    ## render changepoint plot
    output$changepoint_plot <- renderDygraph({
      
      ## load data
      cpts <- cp()
      ts_dat <- time_series()
      means <- mean_series()
      merged <- cbind(ts_data=ts_dat, means_data=xts(means, time(ts_dat)))
      
      ## plot -- "goldfish" color palette from http://www.colourlovers.com/palette/92095/Giant_Goldfish
      dygraph(merged, xlab="Elapsed Time (Hours : Minutes)", ylab="Running Speed (MPH)") %>% 
        dyShading(from=time(ts_dat[cpts[1]]), to=time(ts_dat[cpts[2]]), 
                  color="rgba(105, 210, 231, 0.40)") %>% 
        dyShading(from=time(ts_dat[cpts[2]]), to=time(ts_dat[cpts[3]]), 
                  color="rgba(167, 219, 216, 0.40)") %>% 
        dyShading(from=time(ts_dat[cpts[3]]), to=time(ts_dat[cpts[4]]), 
                  color="rgba(224, 228, 204, 0.40)") %>% 
        dyShading(from=time(ts_dat[cpts[4]]), to=time(ts_dat[cpts[5]]), 
                  color="rgba(243, 134, 48, 0.20)") %>% 
        dyShading(from=time(ts_dat[cpts[5]]), to=time(ts_dat[cpts[6]]), 
                  color="rgba(250, 105, 0, 0.30)") %>% 
        dyEvent(ts_dat[cpts[2]], color="grey") %>% 
        dyEvent(ts_dat[cpts[3]], color="grey") %>% 
        dyEvent(ts_dat[cpts[4]], color="grey") %>% 
        dyEvent(ts_dat[cpts[5]], color="grey") %>% 
        dyAxis("y", valueRange = c(0, 10)) %>%
        dySeries("means_data") %>%
        dySeries("ts_data", label = "Instantaneous Speed") %>%
        dySeries("means_data", label = "Segment Average Speed") %>%
        dyOptions(colors=c("dodgerblue", "deeppink"))
        ## 1) add controls: less changepoints -- more changepoints
        ##                  min segment length: short -- long
    })
    
    ## render cp table
    output$changepoint_table <- renderFormattable({
      formattable(cp_table_data())
    })
})

```

<h4><a href=#calendar-heatmap><i class="fa fa-arrow-left"></i></a> <a href=#trend-analysis><i class="fa fa-arrow-right"></i></a><br>

Changepoint analysis, sometimes called changepoint detection, is a statistical method used to determine if and when there are nominal changes to the values of a time series. Such techniques are commonly used to determine deviations in manufacturing processes, medical device measurements, and website traffic.

In this example, I'm using changepoint analysis to determine shifts in the intensity of recorded runs, in terms of instantaneous speed. (While runners are usually more concerned about average pace, looking at their speed in terms of miles per hour makes more sense in this example.) Each segment is displayed visually, and the associated table details summary statistics from each significant shift in intensity.</h4>
